Local Type Inference for Polarised System F with Existentials

### 2024-04-30 oopsla24/paper.tex
Edit

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  % {\color{red}CHANGE!!}
  % This paper addresses the challenging problem of type inference for
  % Impredicative System F with existential types, a critical aspect of many
  % programming languages. While System F serves as the basis for type systems in
  % numerous languages, existing type inference techniques for Impredicative
  % System F are undecidable due to the presence of existential ($[[∃]]$) 
  % and polymorphic ($[[∀]]$) types. Consequently, current algorithms are often ad-hoc and
  % sub-optimal. This paper presents novel contributions in the form of a local
  % type inference algorithm for Impredicative System F with existential types.
  % The algorithm introduces innovative techniques, such as a unique combination
  % of unification and anti-unification, a full correctness proof, and the use of
  % control structures inspired by \CBPV. Additionally, the paper discusses a type
  % inference framework that allows the algorithm to be applied to different type
  % systems, offering insights into the under-researched area of impredicative
  % existential type inference. 
\end{abstract}

@@

\section{Introduction}

\input{\genDir/tex/01-introduction.tex}

>>>>

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  {\color{red}Change!}
  Type inference for polymorphic types has been extensively studied, with the
  Damas-Hindley-Milner algorithm being the most well-known approach for handling
  parametric polymorphism. However, supporting impredicative existential types
  has remained an open challenge due to undecidability barriers. This paper
  presents \fexists, a type system extending the \CBPV calculus with
  first-class existential and universal quantifiers that can be instantiated
  impredicatively. We give a local type inference algorithm for \fexists that avoids
  the undecidability issues through a novel combination of unification and
  anti-unification techniques. A key contribution is providing a declarative
  specification of the inference algorithm based on the \fexists type system and
  proving the algorithm's soundness and completeness. We also explore extending
  the inference framework to bidirectional typechecking, different subtyping
  regimes, and bounded polymorphism. This work opens up new avenues for bringing
  the benefits of impredicative existential types to programming languages
  through principled type inference.

  % This paper addresses the challenging problem of type inference for
  % Impredicative System F with existential types, a critical aspect of many
  % programming languages. While System F serves as the basis for type systems in
  % numerous languages, existing type inference techniques for Impredicative
  % System F are undecidable due to the presence of existential ($[[∃]]$) 
  % and polymorphic ($[[∀]]$) types. Consequently, current algorithms are often ad-hoc and
  % sub-optimal. This paper presents novel contributions in the form of a local
  % type inference algorithm for Impredicative System F with existential types.
  % The algorithm introduces innovative techniques, such as a unique combination
  % of unification and anti-unification, a full correctness proof, and the use of
  % control structures inspired by \CBPV. Additionally, the paper discusses a type
  % inference framework that allows the algorithm to be applied to different type
  % systems, offering insights into the under-researched area of impredicative
  % existential type inference. 
\end{abstract}

@@

\section{Introduction}
\input{\genDir/tex/01-introduction.tex}

### 2024-04-30 oopsla24/tex/01-introduction.tex
Edit

That is, in pure polymorphic lambda
calculus---\systemf\cite{girard-system-f,jcr-system-f}---the polymorphic type
$[[∀a.fA]]$ has a big lambda $[[Λa.fe]]$ as an introduction form, and an
explicit type application $[[ fe[fA] ]]$ as an elimination form. This is an
extremely simple and compact type system, whose rules fit on a single page, but
whose semantics are sophisticated enough to model things like parametricity and
representation independence. However, System F by itself is unwieldy as a
programming language. The fact that the universal type $[[∀a.fA]]$ has explicit
eliminations means that programs written using polymorphic types will need to be
stuffed to the gills with type annotations explaining how and when to
instantiate the quantifiers.

@@

The most famous of the algorithms for solving these constraints is the
Damas-Hindley-Milner (HDM) algorithm
\cite{hindley69:principal,milner78:theory,damas82:principal}. The idea is that
type instantiation induces a subtype ordering: the type $[[∀a.fA]]$ is a subtype
of all of its instantiations. So we wish to be free to use the same function
$[[len]]$ at many different types such as 
$[[List Int → Int]]$, $[[List Bool → Int]]$,
$[[List (Int × Bool) → Int]]$, and so on.
However, the subtype relation is nondeterministic: it tells us that whenever we
see a polymorphic type $[[∀a.fA]]$, we know it is a subtype of \emph{any}
of its instantiations. To turn this into an algorithm, we have to actually make
some choices, and DHM works by using \emph{unification}. Whenever we would have
had to introduce a particular concrete type in the specification, the DHM
algorithm introduces a unification variable, and incrementally instantiates this
variable as more and more type constraints are observed.

@@

There has been a limited amount of work on support for existential types as a
first-class concept. In an unpublished draft, \citet{leijen06:first-class}
studied an extension of Damas-Milner inference in which type schemes contain
alternating universal and existential quantifiers. Quantifiers still range over
monotypes, and higher-rank polymorphism is not permitted. More recently,
\citet{dunfield16:existential} studied type inference for existential types in
the context of GADT type inference, which, while still predicative, supported
higher-rank types (i.e., quantifiers can occur anywhere inside of a type
scheme). \citet{eisenberg21:existential} propose a system which only permits
types of the form forall-exists, but which permits projective elimination in the
style of ML modules.

All of these papers are restricted to \emph{predicative} quantification, where
quantifiers can only be instantiated with monotypes (i.e., types without any
occurences of quantifiers). However, existential types in full System F are
\emph{impredicative}---that is, quantifiers can be instantiated with arbitrary
types, specifically including types containing quantifiers.

Historically, inference for impredicative quantification has been neglected, due
to results by \citet{tiuryn-urzczyn-96} and \citet{chrzaszcz-98} which show that
not only is full type inference for \systemf undecidable, but even that the
subtyping relation induced by type instantiation is undecidable. However, in
recent years, interest in impredicative inference has revived (for example, the
work of \citet{serrano-2020}), with a focus on avoiding the undecidability
barriers by doing \emph{partial} type inference. That is, we no longer try to do
full type inference, but rather accept that the programmer will need to write
annotations in cases where inference would be too difficult. Unfortunately, it
is often difficult to give a specification for what the partial algorithm
does--for example, \citet{eisenberg21:existential} observe that their algorithm
lacks a declarative specification, and explain why existential types make this
particularly difficult to do.

One especially well-behaved form of partial type inference is \emph{local type
inference}, introduced by \citet{pierce2000:local}. In this approach,
quantifiers in a polymorphic function are instantiated using only the type
information available in the arguments at each call site. While this infers
fewer quantifiers than global algorithms such as Damas-Milner can, it has the
twin benefits that it is easy to give a mathematical specification to, and that
failures of type inference are easily explained in terms of local features of
the call site. In fact, many production programming languages such as C\# and
Java use a form of local type inference, due to the ease of extending this style
of inference to support OO features.

In this paper, we extend the local type inference algorithm to a language with
both universal and existential quantifiers, which can both be instantiated
impredicatively. This combination of features broke a number of the invariants
which traditional type inference algorithms depend on, and required us to invent
new algorithms which combine both unification and (surprisingly)
anti-unification.

@@

To research the type inference in impredicative polymorphic systems
in the presence of existential types, we make the following \emph{contributions}.

\paragraph{Type Inference Algorithm} We give a local type inference algorithm
    which supports both first-class existential and universal quantifiers, both
    of which can be instantiated impredicatively. To evade the undecidability
    results surrounding type inference for \systemf, we introduce \fexists---a
    variant of \CBPV \cite{levy2006:cbpv}, which lets us formulate a subtyping
    relation which is still decidable.

\paragraph{Declarative Specification}
    We give a declarative type system (again, based on \CBPV) to
    serve as a specification of our algorithm, and we prove our algorithm is
    sound and complete with respect to the declarative type system. The
    specification makes it easy to see that all type applications (for
    $\forall$-elimination) and all packs (for $\exists$-introdution) are
    inferred.

\paragraph{Employment of the Anti-unification}
    The introduction of impredicative existentials breaks some of the
    fundamental invariants of HM-style type inference. As a result, it needs to
    mix unification with \emph{anti-unification}---a dual operation which has
    been used in the literature on term rewriting systems, but which has never
    appeared in the context of polymorphic type inference.

\paragraph{Extrapolation of the Inference Framework to Other Type Systems}
    We explore the design space further to show how the
    same type inference scheme could be applied to
    work with different type systems. In particular,
    \begin{enumerate*}
      \item[(i)] in combination with \emph{bidirectional} typechecking---
        the approach used in the original local type inference paper of \citet{pierce2000:local}
        that minimizes the number of needed annotations;
      \item[(ii)] in systems with different subtyping rules, such as  
        the one used in \citet{zhao22:elementary}---a system that
        permits explicit type applications;
      \item[(iii)] in systems with \emph{bounded} polymorphic quantifiers.
        That is, one can only instantiate a quantifier with a 
        type that is a subtype of a given bound.
    \end{enumerate*}

>>>>

That is, in pure polymorphic lambda
calculus---\systemf\cite{girard-system-f,jcr-system-f}---the polymorphic type
$[[∀a.fA]]$ has a big lambda $[[Λa.fe]]$ as an introduction form, and an
explicit type application $[[ fe[fA] ]]$ as an elimination form. This is an
extremely simple and compact type system, whose rules fit on a single page, but
whose semantics are advanced enough to accurately represent concepts such as
parametricity and representation independence. However, \systemf by itself is
unwieldy as a programming language. The fact that the universal type $[[∀a.fA]]$
has explicit eliminations means that programs written using polymorphic types
will need to be stuffed to the gills with type annotations explaining how and
when to instantiate the quantifiers.

@@

The way this is typically done is by using a \emph{subtyping} relation, induced
by the polymorphic instantiation: the type $[[∀a.fA]]$ is a \emph{subtype} of
all of its instantiations. So we wish to be free to use the same polymorphic
function $[[len]]$ at many different types such as $[[List Int → Int]]$, $[[List
Bool → Int]]$, $[[List Int × Bool → Int]]$, and so on. However, the subtyping
can be used nondeterministically: whenever we see a polymorphic type
$[[∀a.fA]]$, we know it is a subtype of \emph{any} of its instantiations. To
turn this into an algorithm, we have to actually make some choices.

The most famous of the algorithms for handling this is the Damas-Hindley-Milner
(DHM) algorithm \cite{hindley69:principal,milner78:theory,damas82:principal}.
DHM deals with this choice using \emph{unification}. Whenever we would have had
to introduce a particular concrete type in the specification, the DHM algorithm
introduces a \emph{unification variable}, and incrementally instantiates this variable
as more and more type constraints are observed.

@@

There has been a limited amount of work on support for existential types as a
first-class concept. In an unpublished draft, \citet{leijen06:first-class}
studied an extension of Damas-Milner inference in which type schemes contain
alternating universal and existential quantifiers. Quantifiers still range over
monotypes (\ie types with no quantifiers inside of them), and higher-rank 
polymorphism is not permitted: one cannot instantiate a quantifier with a type
that itself contains quantifiers.
More recently,
\citet{dunfield16:existential} studied type inference for existential types in
the context of GADT type inference, which, while still predicative, supported
higher-rank types (i.e., quantifiers can occur anywhere inside of a type
scheme). \citet{eisenberg21:existential} propose a system which only permits
types of the form forall-exists, but which permits projective elimination in the
style of ML modules.

All of these papers are restricted to \emph{predicative} quantification, where
quantifiers can only be instantiated with monotypes. However, existential types
in full System F are \emph{impredicative}---that is, quantifiers can be
instantiated with arbitrary types, specifically including types containing
quantifiers.

Historically, inference for impredicative quantification has been neglected, due
to results by \citet{tiuryn-urzczyn-96} and \citet{chrzaszcz-98} which show that
not only is full type inference for \systemf undecidable, but even that the
subtyping relation induced by type instantiation is undecidable. However, in
recent years, interest in impredicative inference has revived (for example, the
work of \citet{serrano-2020}), with a focus on avoiding the undecidability
barriers by doing \emph{partial} type inference. That is, we no longer try to
infer the type of any term in the language, but rather accept that the
programmer will need to write annotations in cases where inference would be too
difficult. Unfortunately, it is often difficult to give a specification for when
the partial algorithm succeeds---for example, \citet{eisenberg21:existential}
observe that their algorithm lacks a declarative specification, and explain why
existential types make this particularly difficult to do.

One especially well-behaved form of partial type inference is \emph{local type
inference}, introduced by \citet{pierce2000:local}. In this approach,
quantifiers in a polymorphic function are instantiated using only the type
information available in the arguments at each call site. While this infers
fewer quantifiers than global algorithms such as Damas-Milner can, it has the
twin benefits that it is easy to give a mathematical specification to, and that
failures of type inference are easily explained in terms of local features of
the call site. In fact, many production programming languages such as C\# and
Java use a form of local type inference, due to the ease of extending this style
of inference to support object-oriented features.

In this paper, we extend the local type inference algorithm to a language with
both universal and existential quantifiers, which can both be instantiated
impredicatively. This combination of features broke a number of the invariants
which traditional type inference algorithms depend on, and required us to invent
new algorithms which combine both unification and (surprisingly)
the dual---anti-unification.

@@

To research the type inference in impredicative polymorphic systems
in the presence of existential types, we make the following \emph{\textbf{contributions}}.

\paragraph{Type Inference Algorithm (\cref{sec:algorithm})}
    We give a local type inference algorithm
    which supports both first-class existential and universal quantifiers, both
    of which can be instantiated impredicatively. To evade the undecidability
    results surrounding type inference for \systemf, we introduce \fexists---a
    variant of \CBPV \cite{levy2006:cbpv}, which lets us formulate a subtyping
    relation which is still decidable.

\paragraph{Declarative Specification (\cref{sec:declarative-system,sec:proofs})}
    We give a declarative type system (again, based on \CBPV) to
    serve as a specification of our algorithm, and we prove our algorithm is
    sound and complete with respect to the declarative type system. The
    specification makes it easy to see that all type applications (for
    $\forall$-elimination) and all packs (for $\exists$-introdution) are
    inferred.

\paragraph{Employment of the Anti-unification (\cref{sec:antiunification})}
    The introduction of impredicative existentials breaks some of the
    fundamental invariants of HM-style type inference. As a result, it needs to
    mix unification with \emph{anti-unification}---a dual operation which has
    been used in the literature on term rewriting systems, but which has never
    appeared in the context of polymorphic type inference.

\paragraph{Extrapolation of the Inference Framework to Other Type Systems
          (\cref{sec:extensions})}
    We explore the design space further to show how the
    same type inference scheme can be applied to
    work with different type systems. In particular,
    \begin{enumerate*}
      \item[(i)] in combination with \emph{bidirectional} typechecking---the 
        approach used in the original local type inference paper of 
        \citet{pierce2000:local} that minimizes the number of needed annotations;
      \item[(ii)] in systems with different subtyping rules, such as  
        the one used in \citet{zhao22:elementary}---a system that
        permits explicit type applications;
      \item[(iii)] in systems with \emph{bounded} polymorphic quantifiers.
        That is, one can only instantiate a quantifier with a 
        type that is a subtype of a given bound.
    \end{enumerate*}

\ilyam{Is the following needed since we have the contributions? I just wanted to introduce the appendix somewhere...}
\paragraph{\textbf{Structure of this Paper}}
In \cref{sec:overview}, we discuss the language of the \fexists system, providing
examples, outlining limitations, and summarizing the main ideas behind the
algorithm. \Cref{sec:declarative-system} presents a formal definition of the
declarative specification and its key properties. In \cref{sec:algorithm}, the
type inference algorithm is introduced, detailing a hierarchy of subroutines;
upon perusing this section, one could \emph{implement} the type inference
algorithm. \Cref{sec:proofs} overviews the algorithm's correctness proof,
explaining its structure and how the declarative system's features make it
decidable. \Cref{sec:extensions} covers potential \fexists extensions and future
work directions. Finally, \cref{appendix} contains comprehensive formal
definitions and proofs of lemmas and theorems used in the paper; in instances
where the full definitions might not be of immediate interest, we direct the
reader to the appendix.

% \paragraph{\textbf{Acknowledgments}} We would like to thank Meven
% Lennon-Bertrand for his extremely helpful comments and suggestions.

### 2024-04-30 oopsla24/tex/02-overview.tex
Edit

% Imprediicative
% CBPV, monadic control
% Local Inference
% Two ways to translate

The language of \fexists is different from standard \systemf:
every term and type has either positive or negative polarity.
The type variables are annotated with their polarities
(e.g. $[[α⁺]]$ or $[[β⁻]]$), and the types can change polarity
via the shift operators $[[↑]]$ (positive to negative) and
$[[↓]]$ (negative to positive).

@@

\paragraph{Non-example: Polymorphic Arguments}
The focus of our system is the \emph{local} type inference
\cite{pierce2000:local}, which has certain limitations. In particular, the
instantiation does not happen when the polymorphic term is in the argument
position. For example, one would expect the following to be inferrable:
\mbox{$[[· ; Φ ⊢ let x = map(id, [two, three, nine]); return x : ↑ List Int]]$}.
However, to infer this, $[[id]]$ must be instantiated to \mbox{$[[Int → ↑Int]]$}
which requires the information to be propagated from the neighboring branch of
the syntax tree (\mbox{$[[ [two, three, nine] ]]$}), \ie bypass the locality.

@@

\begin{itemize}
\item [$\pm$] $[[na]]$ and $[[pa]]$
  denote negative and positive type variables, respectively.
  The variables are either introduced by the corresponding quantifiers ($[[∀]]$ and $[[∃]]$)
  or declared in the type context;
\item [$-$] a function $[[iP → iN]]$ takes a value as input and returns a computation; 
\item [$-$] similarly to functions, a polymorphic abstraction $[[∀pas.iN]]$ quantifies a computation over
  a list of positive type variables $[[pas]]$;
\item [$+$] dually, $[[∃nas.iP]]$, binds negative variables in a positive type $[[iP]]$;
\item [$-$] an upshift $[[↑iP]]$ allows a value to be used as a computation, 
  which at the term level corresponds to a pure computation $[[return v]]$;
\item [$+$] dually, a downshift $[[↓iN]]$ thunks a computation, which at the
  term level corresponds to $[[ {c} ]]$.
\end{itemize}

\paragraph{Definitional Equalities}
For simplicity, we assume that alpha-equivalent terms are equal.
This way, we assume that substitutions do not capture bound variables.
Besides, we assume the quantification is associative by
(congruently) equating $[[∀pas.∀pbs.iN]]$ with $[[∀pas,pbs.iN]]$, 
as well as $[[∃nas.∃nbs.iP]]$ with $[[∃nas,nbs.iP]]$.

@@

In the construction of \fexists, a type context (denoted as $[[Γ]]$) is
represented as a \emph{set} of positive and negative type variables and it is used to
assert the well-formedness of types. The well-formedness of a type is denoted as
$[[Γ ⊢ iP]]$ and $[[Γ ⊢ iN]]$ and it asserts that all type variables are either
bound by a quantifier ($[[∀]]$ and $[[∃]]$) or declared in the context $[[Γ]]$.
The well-formedness checking is an \emph{algorithmic} procedure. As commonly
done, we represent it as a system of inference rules, that correspond to a
recursive algorithm taking the context and the type as input.

@@

In \cref{fig:declarative-terms}, we define the language of terms of 
\fexists. The language combines \systemf with the \CBPV approach.

\begin{itemize}
    \item [$+$] $[[x]]$ denotes a term variable.
      Following the \CBPV stratification, we only have \emph{positive} (value)
      term variables;
    \item [$+$] $[[{c}]]$ is a value corresponding to a thunked 
        or suspended computation;
    \item [$\pm$] $[[(c : iN)]]$ and $[[(v : iP)]]$ allow one to annotate 
        positive and negative terms;
    \item [$-$] $[[return v]]$ is a pure computation, returning a value;
    \item [$-$] $[[λ x : iP . c]]$ and $[[Λ pa . c]]$
        are standard lambda abstractions. Notice that we require
        the type annotation for the argument of $[[λ]]$;
    \item [$-$] $[[ let x = v ; c]]$ is a standard let, binding
        a value $[[v]]$ to a variable $[[x]]$ in a computation $[[c]]$;
    \item [$-$] computational let form $[[ let x : iP = c; c']]$ 
      operates similarly to a monadic bind.
      It binds the result of a computation $[[c : ↑iP]]$ 
      to a variable $[[x : iP]]$, and continues with a computation $[[c']]$;
    \item [$-$] applicative let forms $[[let x : iP = v ( args ) ; c]]$ and
        $[[let x = v ( args ) ; c]]$ can be viewed as a special case of the
        computational let, when the first computation is a function application
        $[[v]] ([[args]])$.
        They take a suspended computation $[[v]]$, apply it to a list of
        arguments, bind the result (which is expected to be pure) to a variable
        $[[x]]$, and continue with a computation $[[c]]$. If the resulting type
        of the application is unique, one can omit the type annotation, as in
        the second form: it will be inferred by the algorithm;
    \item [$-$] $[[let∃ ( nas , x ) = v ; c]]$
        is the standard unpack of an existential type:
        expecting $[[v]]$ to be an existential type,
        it binds the packed negative types to a list of 
        variables $[[nas]]$, binds the body of the existential
        to $[[x]]$, and continues with a computation $[[c]]$.
\end{itemize}

@@

\subsection{The key ideas of the algorithm}

The inference algorithm for \fexists is \emph{local}: it has a limited scope of
the inference information and does not propagate it between far-apart branches
of the syntax tree. Nevertheless, many difficulties appear in the inference of
polymorphic and existential types. Next, we discuss the most challenging of
these difficulties and how they are addressed in the algorithm.

\paragraph{Subtyping} Most of the difficulty of the inference algorithm is
concentrated in the subtyping relation. This is the part of the system where
most of the polymorphic instantiation happens. In particular, it is the
subtyping relation that allows us to use the polymorphic terms in the context
where their concrete instantiations are expected.

@@

\paragraph{Equivalence}
Our system introduces another complexity: the equivalence induced by subtyping
($[[Γ ⊢ iN ≈ iM]] \defiff [[Γ ⊢ iN ≤ iM]] \text{ and } [[Γ ⊢ iM ≤ iN]]$) is not
straightforward. This relation extends beyond just alpha-equivalent types. As
the neighboring polymorphic quantifiers can be instantiated in arbitrary order,
the equivalence relation must permit the permutations of the quantifiers. For
example, we have $[[· ⊢ ∀α⁺.∀β⁺.iN ≈ ∀β⁺.∀α⁺.iN]]$. Similarly, the
equivalence permits the removal/addition of the unused quantifiers: $[[· ⊢
∀α⁺.↑Int ≈ ↑Int]]$.

In order to manage the complexity of non-trivial equivalence, we implement
a type normalization algorithm. It effectively reduces the mutual subtyping to
a simpler alpha-equivalence. The normalization works by recursively
eliminating unused quantifiers and reordering the remaining ones into a
canonical permutation.

@@

To maintain the decidability in the presence of impredicativity, we carefully
restrict the system using the polarity stratification. One of the main
constraints that we put on subtyping is the invariance of the shift operators:
the subtyping $[[↑iP ≤ ↑iQ]]$ is only allowed if $[[iQ ≥ iP]]$
\emph{and} $[[iP ≥ iQ]]$. These constraints do not trivialize the system: we
still have non-trivial upper bounds, and thus, need to use anti-unification to
find them. However, it prevents such examples as $[[α̂⁺ :≥ ↓↑α̂⁺]]$, and in
fact, makes the constraint resolution decidable.

>>>>

\label{sec:overview}

In the presence of impredicative polymorphism, the standard \systemf has
undecidable subtyping and type inference. It means that any inference algorithm
is incomplete: it only infers the types for a \emph{fragment} of \systemf.

To tackle this issue, we restrict \systemf by employing the ideas from the \CBPV
system \cite{levy2006:cbpv}. We introduce a new language, \fexists, which
stratifies the syntax of \systemf into positive and negative parts. It provides
the structure that allows us to restrict the typing specification to make it
decidable.

Every term and type fo \fexists has either \emph{positive} or \emph{negative}
polarity. The type variables are annotated with their polarities (e.g. $[[α⁺]]$
or $[[β⁻]]$), and the types can change polarity via the shift operators $[[↑]]$
(positive to negative) and $[[↓]]$ (negative to positive).

@@

\paragraph{Non-example: Polymorphic Arguments}
The focus of our system is the \emph{local} type inference
\cite{pierce2000:local}, which has certain limitations. In particular, the
instantiation does not happen when the polymorphic term is in the argument
position. For example, one could expect the following to be inferrable:
\mbox{$[[· ; Φ ⊢ let x = map(id, [two, three, nine]); return x : ↑ List Int]]$}.
However, to infer this, $[[id]]$ must be instantiated to \mbox{$[[Int → ↑Int]]$}
which requires the information to be propagated from the neighboring branch of
the syntax tree (\mbox{$[[ [two, three, nine] ]]$}), \ie bypass the locality.

@@

\begin{itemize}
\item [$\pm$] $[[na]]$ and $[[pa]]$
  denote negative and positive type variables, respectively.
  The variables are either introduced by the corresponding quantifiers 
  ($[[∀]]$ and $[[∃]]$, respectively) or declared in the type context;
\item [$-$] a function $[[iP → iN]]$ takes a \emph{positive} input $[[iP]]$ and
returns a \emph{negative} output $[[iN]]$; the function type itself is
\emph{negative}, so the `arrow' operator is right-associative: 
$[[iP → iQ → iN]]$ is equivalent to $[[iP → (iQ → iN)]]$;
\item [$-$] similarly to functions, a polymorphic abstraction $[[∀pas.iN]]$
  quantifies a negative type $[[iN]]$ over a list of positive type variables $[[pas]]$;
\item [$+$] dually, $[[∃nas.iP]]$ binds negative variables in a positive type $[[iP]]$;
\item [$\pm$] an upshift $[[↑iP]]$ and a downshift $[[↓iN]]$ change the polarity of a type
  from positive to negative and vice versa.
  At the level of terms, the constructors for shift types are
  $[[return v]]$ and $[[{c}]]$ (thunked computation), respectively.
\end{itemize}

\paragraph{Syntactic Conventions}
We always consider terms and types up to alpha-equivalence, and capture-avoiding
substitution. Besides, we assume the quantification is associative,
for example, $[[∀pas.∀pbs.iN]]$ and $[[∀pas,pbs.iN]]$
represent the same type.

@@

In the construction of \fexists, a type context (denoted as $[[Γ]]$) is
represented as a \emph{set} of positive and negative type variables and it is
used to assert the well-formedness of types. The well-formedness (or
well-scopeness) of a type is denoted as $[[Γ ⊢ iP]]$ and $[[Γ ⊢ iN]]$ and it
asserts that all type variables are either bound by a quantifier ($[[∀]]$ and
$[[∃]]$) or declared in the context $[[Γ]]$. The well-formedness checking is an
\emph{algorithmic} procedure. We represent it as a system of inference rules,
that correspond to a recursive algorithm taking the context and the type as
input. For brevity, we omit these rules, as they are standard binder scoping
rules.

@@

In \cref{fig:declarative-terms}, we define the language of terms of \fexists.
The language is a combination of \systemf and \CBPV constructs: the
terms are stratified into values and computations; there are return 
and thunk constructors, as well as several forms of let bindings to
sequentialize the computation/application with or without type annotation;

\begin{itemize}
    \item [$+$] $[[x]]$ denotes a term variable.
      Following the \CBPV stratification, we only have \emph{positive} (value)
      term variables;
    \item [$+$] $[[{c}]]$ is a value corresponding to a thunked 
        or suspended computation;
    \item [$\pm$] $[[(c : iN)]]$ and $[[(v : iP)]]$ allow one to annotate 
        positive and negative terms;
    \item [$-$] $[[return v]]$ is a pure computation, returning a value;
    \item [$-$] $[[λ x : iP . c]]$ and $[[Λ pa . c]]$
        are standard lambda abstractions. Notice that we require
        the type annotation for the argument of $[[λ]]$;
    \item [$-$] $[[ let x = v ; c]]$ is a standard let, binding
        a value $[[v]]$ to a variable $[[x]]$ in a computation $[[c]]$;
    \item [$-$] computational let form $[[ let x : iP = c; c']]$ 
      operates similarly to a monadic bind.
      It binds the result of a computation $[[c : ↑iP]]$ 
      to a variable $[[x : iP]]$, and continues with a computation $[[c']]$;
    \item [$-$] applicative let forms $[[let x : iP = v ( args ) ; c]]$ and
        $[[let x = v ( args ) ; c]]$ can be viewed as a special case of the
        computational let, when the first computation is a function application
        $[[v]] ([[args]])$.
        They take a suspended computation $[[v]]$, apply it to a list of
        arguments, bind the (pure) result to a variable
        $[[x]]$, and continues with a computation $[[c]]$. If the resulting type
        of the application is unique, one can omit the type annotation, as in
        the second form: it will be inferred by the algorithm;
    \item [$-$] $[[let∃ ( nas , x ) = v ; c]]$
        is the standard eliminator of an existential type (unpack):
        expecting $[[v]]$ to be of an existential type,
        it binds the packed negative types to a list of 
        variables $[[nas]]$, binds the body of the existential
        to $[[x]]$, and continues with a computation $[[c]]$.
\end{itemize}

@@

\subsection{The Key Ideas of the Algorithm}

The inference algorithm for \fexists is \emph{local}: it has a limited scope of
the inference information and does not propagate it between far-apart branches
of the syntax tree. Nevertheless, many difficulties appear in the inference of
the combination of polymorphic and existential types. Next, we discuss the most
challenging of these difficulties and how they are addressed in the algorithm.

\paragraph{Subtyping} 
The inference algorithm's complexity mainly lies in the subtyping relation,
which is, in particular, responsible for polymorphic instantiation. It's this
subtyping relation that enables us to apply polymorphic terms in situations
where their specific instantiations are anticipated.

@@

\paragraph{Equivalence}
The system introduces another complexity: the equivalence induced by subtyping
($[[Γ ⊢ iN ≈ iM]] \defiff [[Γ ⊢ iN ≤ iM]] \text{ and } [[Γ ⊢ iM ≤ iN]]$) is not
straightforward. This relation extends beyond just alpha-equivalent types. As
the neighboring polymorphic quantifiers can be instantiated in arbitrary order,
the equivalence relation must permit the \emph{permutations} of the quantifiers.
For example, we have $[[· ⊢ ∀α⁺.∀β⁺.iN ≈ ∀β⁺.∀α⁺.iN]]$. Similarly, the
equivalence permits the \emph{removal/addition} of the unused quantifiers: $[[·
⊢ ∀α⁺.↑Int ≈ ↑Int]]$.

To handle the complexity of non-trivial equivalence, we employ a type
normalization algorithm. This approach simplifies mutual subtyping to a more
manageable alpha-equivalence by eliminating unused quantifiers and reordering
the rest into a canonical permutation, effectively replacing subtyping-induced
equivalence on terms with alpha-equivalence on their normal forms.

@@

To maintain the decidability in the presence of impredicativity, we carefully
restrict the system using the polarity stratification. One of the main
constraints that we put on subtyping is the invariance of the shift operators:
the subtyping $[[↑iP ≤ ↑iQ]]$ is only allowed if $[[iQ ≥ iP]]$
\emph{and} $[[iP ≥ iQ]]$. These constraints do not trivialize the system: we
still have non-trivial upper bounds, and thus, need to use anti-unification to
find them. However, it prevents such examples as $[[α̂⁺ :≥ ↓↑α̂⁺]]$, and 
as we prove, makes the constraint resolution decidable.
